### 2021-10-14 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_ncbitaxonomy_softwarenote@d3cea5b5fa2df44d44b68a798cacd1e9358e28d6 🚀

\fancypagestyle{normal}
{
  \fancyhf{}
  \fancyfoot[R]{\footnotesize\sffamily\thepage\ of \pageref*{LastPage}}
}
\begin{document}
\raggedright
\thispagestyle{empty}
{\Large\bfseries\sffamily Template to prepare preprints and manuscripts
using markdown and github actions}
\vskip 5em

%
\href{https://orcid.org/0000-0002-0735-5184}{Timothée\,Poisot}%
%
\,\textsuperscript{1,2,‡}\quad %
Peregrin\,Took%
%
\,\textsuperscript{3,4}\quad %
Merriadoc\,Brandybuck%
%
\,\textsuperscript{5,4,‡}

\textsuperscript{1}\,Université de
Montréal\quad \textsuperscript{2}\,Québec Centre for Biodiversity
Sciences\quad \textsuperscript{3}\,Inn of the Prancing
Pony\quad \textsuperscript{4}\,Fellowship of the
Ring\quad \textsuperscript{5}\,Green Dragon Inn

\textsuperscript{‡}\,These authors contributed equally to the work\\

@@

        {\bfseries Purpose:}\,This template provides a series of scripts
to render a markdown document into an interactive website and a series
of PDFs.\\%
        {\bfseries Motivation:}\,It makes collaborating on text with
GitHub easier, and means that we never need to think about the
output.\\%
        {\bfseries Internals:}\,GitHub actions and a series of python
scritpts. The markdown is handled with \texttt{pandoc}.\\%

@@

\texttt{NCBITaxonomy.jl} is a package designed to facilitate the
reconciliation and cleaning of taxonomic names, using a local copy of
the NCBI taxonomic backbone (Federhen 2012; Schoch \emph{et al.} 2020);
The basic search functions are coupled with quality-of-life functions
including case-insensitive search and custom fuzzy string matching to
facilitate the amount of information that can be extracted automatically
while allowing efficient manual curation and inspection of results.
\texttt{NCBITaxonomy.jl} works with version 1.6 of the Julia programming
language (Bezanson \emph{et al.} 2017), and relies on the Apache Arrow
format to store a local copy of the NCBI raw taxonomy files. The design
of \texttt{NCBITaxonomy.jl} has been inspired by similar efforts, like
the R package \texttt{taxadb} (Norman \emph{et al.} 2020), which
provides an offline alternative to packages like \texttt{taxize}
(Chamberlain \& Szöcs 2013).

@@

Unambiguously identifying species is a far more challenging task than it
may appear. There are a vast number of reasons for this. Different
databases keep different taxonomic ``backbones,'' \emph{i.e.} different
data structures in which names are mapped to species, and organised in a
hierarchy. Not all names are unique identifiers to groups. For example,
\emph{Io} can either refer to a genus of plants from the aster family,
or to a genus of molluscs; the genus \emph{Mus} (of which the house
mouse \emph{Mus musculus} is a species), contains a sub-genus
\emph{also} named \emph{Mus}. Conversely, the same species can have
several names, which are valid synonyms: for example, the domestic cow
\emph{Bos taurus} admits \emph{Bos primigenius taurus} as a valid
synonym. Taxonomic nomenclature also changes regularly, with groups
being split, merged, or moved to a new position in the tree of life;
this is, notably, a common occurrence with viral taxonomy, each
subsequent version of which can differ markedly from the last; compare,
\emph{e.g} Lefkowitz \emph{et al.} (2018) to Walker \emph{et al.}
(2020).

To add to the complexity, one must also consider that most taxa names
are at some point manually typed, which has the potential to introduce
additional mistakes in raw data; it is likely to expect that such
mistakes may arise when attempting to write down the (perfectly valid)
names of the bacterial isolate known as \emph{Myxococcus
llanfairpwllgwyngyllgogerychwyrndrobwllllantysiliogogogochensis}, or of
the crowned slaty flycatcher \emph{Griseotyrannus
aurantioatrocristatus}. These mistakes are more likely when dealing with
hyper-diverse samples, like plant census (Dauncey \emph{et al.} 2016;
Wagner 2016; Conti \emph{et al.} 2021). In addition to binomial names,
the same species can be known by many vernacular (common) names, which
are language or even region-specific: \emph{Ovis aries}, for example,
has valid English vernaculars including lamb, sheep, wild sheep, and
domestic sheep.

@@

A package with the ability to handle the sources of errors outlined
above, and especially while provide an authoritative classification, can
accelerate the work of consuming large volumes of biodiversity data. For
example, this package was used in the process of developing the
\emph{CLOVER} database (Gibb \emph{et al.} 2021) of host-virus
associations, by reconciling the names of viruses and mammals from four
different sources, where all of the issues described above were present.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-Bezanson2017JulFre}{}%
Bezanson, J., Edelman, A., Karpinski, S. \& Shah, V. (2017). Julia: A
Fresh Approach to Numerical Computing. \emph{SIAM Review}, 59, 65--98.

\leavevmode\hypertarget{ref-Chamberlain2013TaxTax}{}%
Chamberlain, S.A. \& Szöcs, E. (2013). Taxize: Taxonomic search and
retrieval in R. \emph{F1000Research}, 2, 191.

\leavevmode\hypertarget{ref-Conti2021MatAlg}{}%
Conti, M., Nimis, P.L. \& Martellos, S. (2021). Match Algorithms for
Scientific Names in FlorItaly, the Portal to the Flora of Italy.
\emph{Plants}, 10, 974.

\leavevmode\hypertarget{ref-Dauncey2016ComMis}{}%
Dauncey, E.A., Irving, J., Allkin, R. \& Robinson, N. (2016). Common
mistakes when using plant names and how to avoid them. \emph{European
Journal of Integrative Medicine}, 8, 597--601.

\leavevmode\hypertarget{ref-Federhen2012NcbTax}{}%
Federhen, S. (2012). The NCBI taxonomy database. \emph{Nucleic acids
research}, 40, D136--D143.

\leavevmode\hypertarget{ref-Gibb2021DatPro}{}%
Gibb, R., Albery, G.F., Becker, D.J., Brierley, L., Connor, R., Dallas,
T.A., \emph{et al.} (2021). Data proliferation, reconciliation, and
synthesis in viral ecology. \emph{bioRxiv}, 2021.01.14.426572.

\leavevmode\hypertarget{ref-Lefkowitz2018VirTax}{}%
Lefkowitz, E.J., Dempsey, D.M., Hendrickson, R.C., Orton, R.J., Siddell,
S.G. \& Smith, D.B. (2018). Virus taxonomy: The database of the
International Committee on Taxonomy of Viruses (ICTV). \emph{Nucleic
Acids Research}, 46, D708--D717.

\leavevmode\hypertarget{ref-Norman2020TaxHig}{}%
Norman, K.E.A., Chamberlain, S. \& Boettiger, C. (2020). Taxadb: A
high-performance local taxonomic database interface. \emph{Methods in
Ecology and Evolution}, 11, 1153--1159.

\leavevmode\hypertarget{ref-Schoch2020NcbTax}{}%
Schoch, C.L., Ciufo, S., Domrachev, M., Hotton, C.L., Kannan, S.,
Khovanskaya, R., \emph{et al.} (2020). NCBI Taxonomy: A comprehensive
update on curation, resources and tools. \emph{Database}, 2020.

\leavevmode\hypertarget{ref-Shimatani2001MeaSpe}{}%
Shimatani, K. (2001). On the Measurement of Species Diversity
Incorporating Species Differences. \emph{Oikos}, 93, 135--147.

\leavevmode\hypertarget{ref-Wagner2016RevSof}{}%
Wagner, V. (2016). A review of software tools for spell-checking taxon
names in vegetation databases. \emph{Journal of Vegetation Science}, 27,
1323--1327.

\leavevmode\hypertarget{ref-Walker2020ChaVir}{}%
Walker, P.J., Siddell, S.G., Lefkowitz, E.J., Mushegian, A.R.,
Adriaenssens, E.M., Dempsey, D.M., \emph{et al.} (2020). Changes to
virus taxonomy and the Statutes ratified by the International Committee
on Taxonomy of Viruses (2020). \emph{Archives of Virology}, 165,
2737--2748.

>>>>

\fancypagestyle{normal}
{
  \fancyhf{}
  \fancyfoot[R]{\footnotesize\sffamily\thepage\ of \pageref*{LastPage}}
}
\begin{document}
\raggedright
\thispagestyle{empty}
{\Large\bfseries\sffamily NCBITaxonomy.jl - rapid biological names
finding and reconciliation}
\vskip 5em

%
\href{https://orcid.org/0000-0002-0735-5184}{Timothée\,Poisot}%
%
\,\textsuperscript{1,2}\quad %
\href{https://orcid.org/0000-0002-0965-1649}{Rory\,Gibb}%
%
\,\textsuperscript{3,4}\quad %
\href{https://orcid.org/0000-0002-4308-6321}{Sadie J.\,Ryan}%
%
\,\textsuperscript{5,6,7}\quad %
\href{https://orcid.org/0000-0001-6960-8434}{Colin J.\,Carlson}%
%
\,\textsuperscript{8,9}

\textsuperscript{1}\,Université de Montréal, Départment de Sciences
Biologiques, Montréal QC, Canada\quad \textsuperscript{2}\,Québec Centre
for Biodiversity Science, Montréal, QC,
Canada\quad \textsuperscript{3}\,Centre on Climate Change and Planetary
Health, London School of Hygiene and Tropical Medicine, London,
UK\quad \textsuperscript{4}\,Centre for Mathematical Modelling of
Infectious Diseases, London School of Hygiene and Tropical Medicine,
London, UK\quad \textsuperscript{5}\,Emerging Pathogens Institute,
University of Florida, Gainesville, FL, United States of
America\quad \textsuperscript{6}\,School of Life Sciences, University of
KwaZulu-Natal, Durban, South Africa\quad \textsuperscript{7}\,Department
of Geography, University of Florida, Gainesville, FL, United States of
America\quad \textsuperscript{8}\,Department of Microbiology and
Immunology, Georgetown University Medical Center, Georgetown University,
Washington, D.C., United States of
America\quad \textsuperscript{9}\,Center for Global Health Science and
Security, Georgetown University Medical Center, Georgetown University,
Washington, D.C., United States of America

@@

\vfill
\texttt{NCBITaxonomy.jl} is a package designed to facilitate the
reconciliation and cleaning of taxonomic names, using a local copy of
the NCBI taxonomic backbone (Federhen 2012, Schoch et al. 2020); The
basic search functions are coupled with quality-of-life functions
including case-insensitive search and custom fuzzy string matching to
facilitate the amount of information that can be extracted automatically
while allowing efficient manual curation and inspection of results.
\texttt{NCBITaxonomy.jl} works with version 1.6 of the Julia programming
language (Bezanson et al. 2017), and relies on the Apache Arrow format
to store a local copy of the NCBI raw taxonomy files. The design of
\texttt{NCBITaxonomy.jl} has been inspired by similar efforts, like the
R package \texttt{taxadb} (Norman et al. 2020), which provides an
offline alternative to packages like \texttt{taxize} (Chamberlain and
Szöcs 2013).

@@

\texttt{NCBITaxonomy.jl} is a package designed to facilitate the
reconciliation and cleaning of taxonomic names, using a local copy of
the NCBI taxonomic backbone (Federhen 2012, Schoch et al. 2020); The
basic search functions are coupled with quality-of-life functions
including case-insensitive search and custom fuzzy string matching to
facilitate the amount of information that can be extracted automatically
while allowing efficient manual curation and inspection of results.
\texttt{NCBITaxonomy.jl} works with version 1.6 of the Julia programming
language (Bezanson et al. 2017), and relies on the Apache Arrow format
to store a local copy of the NCBI raw taxonomy files. The design of
\texttt{NCBITaxonomy.jl} has been inspired by similar efforts, like the
R package \texttt{taxadb} (Norman et al. 2020), which provides an
offline alternative to packages like \texttt{taxize} (Chamberlain and
Szöcs 2013).

@@

Unambiguously identifying species is a far more challenging task than it
may appear. There are a vast number of reasons for this. Different
databases keep different taxonomic ``backbones,'' \emph{i.e.} different
data structures in which names are mapped to species, and organised in a
hierarchy. Not all names are unique identifiers to groups. For example,
\emph{Io} can either refer to a genus of plants from the aster family,
or to a genus of molluscs; the genus \emph{Mus} (of which the house
mouse \emph{Mus musculus} is a species), contains a sub-genus
\emph{also} named \emph{Mus}. Conversely, the same species can have
several names, which are valid synonyms: for example, the domestic cow
\emph{Bos taurus} admits \emph{Bos primigenius taurus} as a valid
synonym. Taxonomic nomenclature also changes regularly, with groups
being split, merged, or moved to a new position in the tree of life;
this is, notably, a common occurrence with viral taxonomy, each
subsequent version of which can differ markedly from the last; compare,
\emph{e.g} Lefkowitz et al. (2018) to Walker et al. (2020).

To add to the complexity, one must also consider that most taxa names
are at some point manually typed, which has the potential to introduce
additional mistakes in raw data; it is likely to expect that such
mistakes may arise when attempting to write down the (perfectly valid)
names of the bacterial isolate known as \emph{Myxococcus
llanfairpwllgwyngyllgogerychwyrndrobwllllantysiliogogogochensis}, or of
the crowned slaty flycatcher \emph{Griseotyrannus
aurantioatrocristatus}. These mistakes are more likely when dealing with
hyper-diverse samples, like plant census (Dauncey et al. 2016, Wagner
2016, Conti et al. 2021). In addition to binomial names, the same
species can be known by many vernacular (common) names, which are
language or even region-specific: \emph{Ovis aries}, for example, has
valid English vernaculars including lamb, sheep, wild sheep, and
domestic sheep.

@@

A package with the ability to handle the sources of errors outlined
above, and especially while provide an authoritative classification, can
accelerate the work of consuming large volumes of biodiversity data. For
example, this package was used in the process of developing the
\emph{CLOVER} database (Gibb et al. 2021) of host-virus associations, by
reconciling the names of viruses and mammals from four different
sources, where all of the issues described above were present.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-Bezanson2017JulFre}{}%
Bezanson, J. et al. 2017. Julia: A Fresh Approach to Numerical
Computing. - SIAM Review 59: 65--98.

\leavevmode\hypertarget{ref-Chamberlain2013TaxTax}{}%
Chamberlain, S. A. and Szöcs, E. 2013. Taxize: Taxonomic search and
retrieval in R. - F1000Research 2: 191.

\leavevmode\hypertarget{ref-Conti2021MatAlg}{}%
Conti, M. et al. 2021. Match Algorithms for Scientific Names in
FlorItaly, the Portal to the Flora of Italy. - Plants 10: 974.

\leavevmode\hypertarget{ref-Dauncey2016ComMis}{}%
Dauncey, E. A. et al. 2016. Common mistakes when using plant names and
how to avoid them. - European Journal of Integrative Medicine 8:
597--601.

\leavevmode\hypertarget{ref-Federhen2012NcbTax}{}%
Federhen, S. 2012. The NCBI taxonomy database. - Nucleic acids research
40: D136--D143.

\leavevmode\hypertarget{ref-Gibb2021DatPro}{}%
Gibb, R. et al. 2021. Data proliferation, reconciliation, and synthesis
in viral ecology. - bioRxiv: 2021.01.14.426572.

\leavevmode\hypertarget{ref-Lefkowitz2018VirTax}{}%
Lefkowitz, E. J. et al. 2018. Virus taxonomy: The database of the
International Committee on Taxonomy of Viruses (ICTV). - Nucleic Acids
Research 46: D708--D717.

\leavevmode\hypertarget{ref-Norman2020TaxHig}{}%
Norman, K. E. A. et al. 2020. Taxadb: A high-performance local taxonomic
database interface. - Methods in Ecology and Evolution 11: 1153--1159.

\leavevmode\hypertarget{ref-Schoch2020NcbTax}{}%
Schoch, C. L. et al. 2020. NCBI Taxonomy: A comprehensive update on
curation, resources and tools. - Database in press.

\leavevmode\hypertarget{ref-Shimatani2001MeaSpe}{}%
Shimatani, K. 2001. On the Measurement of Species Diversity
Incorporating Species Differences. - Oikos 93: 135--147.

\leavevmode\hypertarget{ref-Wagner2016RevSof}{}%
Wagner, V. 2016. A review of software tools for spell-checking taxon
names in vegetation databases. - Journal of Vegetation Science 27:
1323--1327.

\leavevmode\hypertarget{ref-Walker2020ChaVir}{}%
Walker, P. J. et al. 2020. Changes to virus taxonomy and the Statutes
ratified by the International Committee on Taxonomy of Viruses (2020). -
Archives of Virology 165: 2737--2748.

### 2021-10-14 preprint.tex
Deploying to gh-pages from @ PoisotLab/ms_ncbitaxonomy_softwarenote@d3cea5b5fa2df44d44b68a798cacd1e9358e28d6 🚀

\noindent {\LARGE \textbf{\textsf{Template to prepare preprints and
manuscripts using markdown and github actions}}}

\medskip
\begin{flushleft}
{\small
%
\href{https://orcid.org/0000-0002-0735-5184}{Timothée\,Poisot}%
%
\,\textsuperscript{1,2,‡}, %
Peregrin\,Took%
%
\,\textsuperscript{3,4}, %
Merriadoc\,Brandybuck%
%
\,\textsuperscript{5,4,‡}
\vskip 1em
\textsuperscript{1}\,Université de Montréal; \textsuperscript{2}\,Québec
Centre for Biodiversity Sciences; \textsuperscript{3}\,Inn of the
Prancing Pony; \textsuperscript{4}\,Fellowship of the
Ring; \textsuperscript{5}\,Green Dragon Inn\\
\textsuperscript{‡}\,These authors contributed equally to the work\\
\vskip 1em
\textbf{Correspondance to:}\\
Timothée Poisot --- \texttt{timothee.poisot@umontreal.ca}\\
}
\end{flushleft}

@@

\marginpar{\vskip 1em\flushright
{\small{\bfseries Keywords}:\par
pandoc\\pandoc-crossref\\github actions\\}
}

        {\bfseries Purpose:}\,This template provides a series of scripts
to render a markdown document into an interactive website and a series
of PDFs.\\%
        {\bfseries Motivation:}\,It makes collaborating on text with
GitHub easier, and means that we never need to think about the
output.\\%
        {\bfseries Internals:}\,GitHub actions and a series of python
scritpts. The markdown is handled with \texttt{pandoc}.\\%

>>>>

\noindent {\LARGE \textbf{\textsf{NCBITaxonomy.jl - rapid biological
names finding and reconciliation}}}

\medskip
\begin{flushleft}
{\small
%
\href{https://orcid.org/0000-0002-0735-5184}{Timothée\,Poisot}%
%
\,\textsuperscript{1,2}, %
\href{https://orcid.org/0000-0002-0965-1649}{Rory\,Gibb}%
%
\,\textsuperscript{3,4}, %
\href{https://orcid.org/0000-0002-4308-6321}{Sadie J.\,Ryan}%
%
\,\textsuperscript{5,6,7}, %
\href{https://orcid.org/0000-0001-6960-8434}{Colin J.\,Carlson}%
%
\,\textsuperscript{8,9}
\vskip 1em
\textsuperscript{1}\,Université de Montréal, Départment de Sciences
Biologiques, Montréal QC, Canada; \textsuperscript{2}\,Québec Centre for
Biodiversity Science, Montréal, QC, Canada; \textsuperscript{3}\,Centre
on Climate Change and Planetary Health, London School of Hygiene and
Tropical Medicine, London, UK; \textsuperscript{4}\,Centre for
Mathematical Modelling of Infectious Diseases, London School of Hygiene
and Tropical Medicine, London, UK; \textsuperscript{5}\,Emerging
Pathogens Institute, University of Florida, Gainesville, FL, United
States of America; \textsuperscript{6}\,School of Life Sciences,
University of KwaZulu-Natal, Durban, South
Africa; \textsuperscript{7}\,Department of Geography, University of
Florida, Gainesville, FL, United States of
America; \textsuperscript{8}\,Department of Microbiology and Immunology,
Georgetown University Medical Center, Georgetown University, Washington,
D.C., United States of America; \textsuperscript{9}\,Center for Global
Health Science and Security, Georgetown University Medical Center,
Georgetown University, Washington, D.C., United States of America\\
\vskip 1em
\textbf{Correspondance to:}\\
Timothée Poisot --- \texttt{timothee.poisot@umontreal.ca}\\
}
\end{flushleft}

@@

\marginpar{\vskip 1em\flushright
{\small{\bfseries Keywords}:\par
biodiversity\\taxonomy\\NCBI\\fuzzy matching\\}
}

\texttt{NCBITaxonomy.jl} is a package designed to facilitate the
reconciliation and cleaning of taxonomic names, using a local copy of
the NCBI taxonomic backbone (Federhen 2012, Schoch et al. 2020); The
basic search functions are coupled with quality-of-life functions
including case-insensitive search and custom fuzzy string matching to
facilitate the amount of information that can be extracted automatically
while allowing efficient manual curation and inspection of results.
\texttt{NCBITaxonomy.jl} works with version 1.6 of the Julia programming
language (Bezanson et al. 2017), and relies on the Apache Arrow format
to store a local copy of the NCBI raw taxonomy files. The design of
\texttt{NCBITaxonomy.jl} has been inspired by similar efforts, like the
R package \texttt{taxadb} (Norman et al. 2020), which provides an
offline alternative to packages like \texttt{taxize} (Chamberlain and
Szöcs 2013).
